Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*

### 2020-06-08 paper.Rmd
call this version 0.1
Ready to share with close friends.

# Review of Blair and Sambanis 2020

@@

# Replication problems

While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation. These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) incorrect implementations of the weighted by PITF and PITF analog split-population models, c) inconsistent test sets for the models examined, and d) incorrect scoring of the 2016 forecasts. We also note (e) additional concerns arising from the question of how the random forest models were tuned by B&S, especially given the way they are used is unorthodox. We did not further investigate the latter issue as it is rendered somewhat moot by the changes in results after addressing the preceding issues.

@@

The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, actual ROC curves. A reference to smoothing is made in a single sentence in the paper (p. 12):

ROC curves typically appear step-like in response to the distribution of positive and negative cases in the data, which specifically in rare events problems like this one restricts the number of distinct true positive rate values. The specific predictions here also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines as seen in the panels on the right. In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for a observed positive case is reached. For these ROC curves, and for that matter in the basic train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set. Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates.  Notice also that the smoothing averages the left most almost straight line with the right most almost straight line in a monotonic way. This ignores the fact that the actual ROC **MIKE**

As we show below, the use of smoothed ROC curves to calculuate AUC has a dramatic impact on the results. To our knowledge, the norm is to calculate AUC-ROC values using original, not smoothed ROC curves. We are in fact not aware of other work that uses smoothed ROC curves for AUC calculations. It might be that there are theoretical reasons justifying the use of smoothed ROC curves over the original ROC curves; but given that this decision dramatically impacts the interpretation of the B&S results, it minimally would have warranted an explicit discussion in the paper. As it is, B&S offer no justification for using smoothed ROC curves in the AUC-ROC calculations.

@@

We infer that the intent is that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. B&S, however, actually weights the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`). The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions. Similarly in the 6-month version of this file.] This appears to be a coding error.

@@

Similarly, the "PITF Split Population" model appears to be incorrectly implemented. B&S describe it on page 20:

@@

The intention clearly was to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they are both run on the *exact same training data*, which consists of the full training data all other models are run on. The models are also identical otherwise, i.e. they use the same *x* variables and the same random forest hyper-parameter settings. The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself. If we ran both with the same random seed, they would be identical in every respect, producing identical predictions.^[Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined. We have documented details at https://github.com/rickmorgan2/Blair-Sambanis-replication/issues/5.]

@@

@@

## Random forest hyperparameters

We now turn to an examination of our replication that fixes the above issues.  The bottom line is that the changes we make turn on their heads the conclusions of B&S.

@@

B&S have a clear misunderstanding of the role that out-of-sample prediction can play in analysis.  On the one hand it can be used for simple forecasting, while on the other it can be used to evaluate the performance of models, specifically to overfitting and bias.  They further misrepresent cross-validation---which is not about fishing for results, as claimed by B&S.  The current state-of-the-art uses prediction and cross-validation most frequently to provide supportive evidence that is independent of the estimation procedure and the in-sample data. It is incorrect to paint this procedure as atheoretical since many studies will have some explanation for how the model was constructed. These procedures may simply be used to provide evidence for a theoretical argument.  Though that argument may be unconvincing and/or atheoretical.

The B&S approach that is advocated is to use theory to guide prediction. But theory is an ambiguous and undefined concept.  It is not a procedure.  What they actually do is to create a model with four right hand side variables that is supposed to capture a complicated repression-dissent dynamic.  The dynamic is probably not linear, but their model is. There is a wide-ranging literature on this dynamic that they do not rely on to construct their model.  As such their baseline comparison is a poor standard bearer for strong theory.

Further, they completely misunderstand the use of ICEWS event data in current research. They claim that most uses to date have focused on the quad categories, but this ignores a wide swath of literature (Steinert-Threlkeld APSR 2017) and Metternich et al (AJPS 2013) that uses a specific action---such as protest---defined in the CAMEO ontology. In the AJPS article we hand coded, for example, every actor in Thailand and focused on an analysis of how those have been interacting.

@@

@@

```
\newpage

# Appendix

@@

>>>>

# Summary of Blair and Sambanis (2020)

@@

# Implementation issues in B&S

While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation. These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) incorrect implementations of the weighted by PITF and PITF analog split-population models, c) inconsistent test sets for the models examined, and d) incorrect scoring of the 2016 forecasts.^[We also note (e) additional concerns arising from the question of how the random forest models were tuned by B&S, especially given the way they are used is unorthodox. We did not further investigate the latter issue as it is rendered somewhat moot by the changes in results after addressing the preceding issues.]

@@

The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, actual ROC curves. the data, which specifically in rare events problems like this one restricts the number of distinct true positive rate values. As we show below, the use of smoothed ROC curves to calculuate AUC has a dramatic impact on the results. To our knowledge, the norm is to calculate AUC-ROC values using original, not smoothed ROC curves. We are in fact not aware of other work that uses smoothed ROC curves for AUC calculations. It might be that there are theoretical reasons justifying the use of smoothed ROC curves over the original ROC curves; but given that this decision dramatically impacts the interpretation of the B&S results, it minimally would have warranted an explicit discussion in the paper. As it is, B&S offer no justification for using smoothed ROC curves in the AUC-ROC calculations.

@@

We infer that the intent is that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. B&S, however, actually weights the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`). The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions. Similarly in the 6-month version of this file.] This appears to be an easily corrected coding error on the part of B&S.

@@

Similarly, the "PITF Split Population" model appears to be incorrectly implemented, owing to a coding error. B&S describe it on page 20:

@@

The intention if B&S appears to have been to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they are both run on the *exact same training data*, which consists of the full training data all other models are run on. The models are also identical otherwise, i.e. they use the same *x* variables and the same random forest hyper-parameter settings. The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself. If we ran both with the same random seed, they would be identical in every respect, producing identical predictions.^[Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined. We have documented details at https://github.com/rickmorgan2/Blair-Sambanis-replication/issues/5.]

@@

We now turn to an examination of our analysis that addresses and fixes the issues discussed above.^[The code for all of our analysis undertaken for this effort may be found at https://github.com/rickmorgan2/Blair-Sambanis-replication.]  The bottom line is that the changes we make turn on their heads the conclusions of B&S.

## Smoothed ROC curves and AUC calculations

A reference to smoothing is made in a single sentence in the paper (p. 12):

The specific predictions here also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines as seen in the panels on the right. In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for a observed positive case is reached. For these ROC curves, and for that matter in the basic train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set. Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates.  Notice also that the smoothing averages the left most almost straight line with the right most almost straight line in a monotonic way.

## Using Inconsistent Ns

@@

@@

The B&S approach that is advocated is to use theory to guide prediction. But theory is an ambiguous and undefined concept.  It is not a procedure.  What they actually do is to create a model with four right hand side variables that is supposed to capture a complicated repression-dissent dynamic.  The dynamic is probably not linear, but their model is. There is a wide-ranging literature on this dynamic that they do not rely on to construct their model.  As such their baseline comparison is a poor standard bearer for strong theory. Moreover, they misunderstand the use of ICEWS event data in current research. They claim that most uses to date have focused on the quad categories, but this ignores a wide swath of literature (Steinert-Threlkeld APSR 2017) and Metternich et al (AJPS 2013) that uses a specific action---such as protest---defined in the CAMEO ontology. In the AJPS article on Thailand, Metternich et al hand coded, for example, every actor in Thailand and focused on an analysis of how those interact.

@@

\newpage

# (APPENDIX) Appendix {-}
# Appendix A

@@

```

@@

# Appendix B
## Some additional, minor issues
### Random forest hyperparameters
